---
title: "TCGA_breast.rmd"
author: "Robbin Nameki"
date: '2022-10-19'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Coding exercise to analyze and explore TCGA breast data. 

Part 0 - reading, cleaning and re-formatting data sets 
#Sourcecode
```{r, eval = TRUE}
source('rscripts/Utils.R')
```

#re-formatting rsem matrix to make a) an annotation file (rsem_annot) and b) associated counts matrix (rsem_long)
```{r, eval = TRUE, error = FALSE, warnings = FALSE}
#reading data, filtering for primary tumors and setting rownames to be patient barcode
rsem <- read_rsem() %>%
  filter(sample_type == "Primary Tumor")
rownames(rsem) <- rsem$bcr_patient_barcode

#saving annotation columns & making sample_id column to add clinical & copy number data later
rsem_annot <- rsem[,c("patient_id",
                      "bcr_patient_barcode",
                      "sample_type")] %>%
  mutate(sample_id = 
            str_trunc(bcr_patient_barcode,
            width = 15,
            side = "right",
            ellipsis = "")) 

#reshaping rsem to long format
rsem_long <- as.data.frame(t(rsem[,c(4:20505)])) 
```

#reading & adding relevant clinical and copy number information to annotation file
```{r, eval = TRUE, error = FALSE, warnings = FALSE}
#reading & wrangling clinical data
clinical_data <- read_clinical_data()%>%
  dplyr::select(Sample.ID,
                IHC.HER2,
                HER2.ihc.score,
                HER2.fish.status) %>%
  dplyr::rename(sample_id = Sample.ID,
                ihc_her2 = IHC.HER2,
                ihc_her2_score =  HER2.ihc.score,
                fish_her2_status = HER2.fish.status) 
#reading copy number data
copy_number <- read_copy_number() %>%
  dplyr::select(!patient_id)

#merging clinical and copy number data to make annotation with all data (rsem_annot_all)
rsem_annot_all <- merge(rsem_annot,
                        clinical_data,
                        by = "sample_id",
                        all.x = TRUE)
rsem_annot_all <- merge(rsem_annot_all,
                        copy_number,
                        by = "sample_id",
                        all.x = TRUE)
rownames(rsem_annot_all) <- rsem_annot_all$bcr_patient_barcode
```

#re-setting column order of counts matrix to be the same as the rownames of annotation file
```{r, eval = TRUE, error = FALSE, warnings = FALSE}
#re-setting columns of matrix
rsem_long <- rsem_long[,rownames(rsem_annot_all)]

#double checking that rows and columns match between the two variables
print(all(rownames(rsem_annot_all) == colnames(rsem_long)))

#removing irrelevant variables up to this point
rm(clinical_data,
   copy_number,
   rsem,
   rsem_annot)
```

Part 1 - QC and normalize the RNA dataset.
#Quality control of raw counts
```{r, eval = TRUE, error = FALSE, warnings = FALSE}
#making DESeqDataSet. Rounding counts matrix to fit with DESeq2 input requirements.
dds <- DESeqDataSetFromMatrix(countData = round(rsem_long),
                              colData = rsem_annot_all,
                              design = ~ sample_type)
#Total sample count
a <- plot_total_counts(dds)

#Number of genes detected based on thresholds
b <- plot_gene_detection(dds)

ggarrange(a,b)
rm(a,b)
```

#Normalizing RNA-seq data set
```{r, eval = TRUE, error = FALSE, warnings = FALSE}
#Option 1 - counts per million. TPM is more ideal, but gene length is not available. 
rsem_long_norm <- rsem_long/(colSums(rsem_long)/1000000)
rsem_long_norm_log <- log2(rsem_long_norm+1)


#Option 2 - vst normalization with DESeq2. Caveat is that rsem data needs to be rounded and is not ideal.
#vst normalizing data
vsd <- vst(dds, 
           blind=FALSE)

```

#2. Identify HER2+ patients as determined clinically.
```{r, eval = TRUE, error = FALSE, warnings = FALSE}
#stacked barchart describing proportion of her2 status patients by IHC
a <- rsem_annot_all %>%
  drop_na(ihc_her2) %>%
  group_by(ihc_her2) %>%
  summarize(n = n()) %>%
  ggplot(aes(fill= ihc_her2, 
             y=n, 
             x="ihc_her2")) + 
    geom_bar(position="fill", 
             stat="identity") +
  theme_classic()

#stacked barchart describing proportion of her2 status patients by FISH
b <- rsem_annot_all %>%
  drop_na(fish_her2_status) %>%
  group_by(fish_her2_status) %>%
  filter(!fish_her2_status == "[Not Evaluated]") %>%
  summarize(n = n()) %>%
  ggplot(aes(fill= fish_her2_status, 
             y=n, 
             x="fish_her2_status")) + 
    geom_bar(position="fill", 
             stat="identity") +
  theme_classic()

ggarrange(a,
          b)
rm(a,b)
```

#3. Cluster the cancer RNA data (PCA, UMAP, etc).
```{r, eval = TRUE, error = FALSE, warnings = FALSE}
set.seed(1)
plot_sample_clustering(vsd, 
                       anno_vars = c("ihc_her2"), 
                       distance = "euclidean")

plot_pca(vsd, 
         PC_x = 1, 
         PC_y = 2, 
         color_by = "ihc_her2")

plot_pca(vsd, 
         PC_x = 1, 
         PC_y = 2, 
         color_by = "fish_her2_status")

plot_pca(vsd, 
         PC_x = 1, 
         PC_y = 2, 
         color_by = "erbb2_copy_number")
```

#4. Analyze how RNA, copy number, and clinical status are correlated,
```{r, eval = TRUE, error = FALSE, warnings = FALSE}




