---
title: "TCGA_breast.rmd"
author: "Robbin Nameki"
date: '2022-10-19'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Coding exercise to analyze and explore TCGA breast data. 

# Reading, cleaning and re-formatting data sets 
## Sourcecode
```{r, eval = TRUE}
source('rscripts/Utils.R')
```

## re-formatting rsem matrix to make a) an annotation file (rsem_annot) and b) associated counts matrix (rsem_long)
```{r, eval = TRUE, error = FALSE, warnings = FALSE}
#reading data and setting rownames to be patient barcode
rsem <- read_rsem() 
rownames(rsem) <- rsem$bcr_patient_barcode

#saving annotation columns & making sample_id column to add clinical & copy number data later
rsem_annot <- rsem[,c("patient_id",
                      "bcr_patient_barcode",
                      "sample_type")] %>%
  mutate(sample_id = 
            str_trunc(bcr_patient_barcode,
            width = 15,
            side = "right",
            ellipsis = "")) 

#reshaping rsem to long format
rsem_long <- as.data.frame(t(rsem[,c(4:20505)])) 
```

## reading & adding relevant HER2/ERBB2 clinical and copy number information to annotation file. Answers Question 2 - "Identify HER2+ patients as determined clinically."
```{r, eval = TRUE, error = FALSE, warnings = FALSE}
#reading & wrangling clinical data
clinical_data <- read_clinical_data()%>%
  dplyr::select(Sample.ID,
                IHC.HER2,
                HER2.ihc.score,
                HER2.fish.status) %>%
  dplyr::rename(sample_id = Sample.ID,
                ihc_her2 = IHC.HER2,
                ihc_her2_score =  HER2.ihc.score,
                fish_her2_status = HER2.fish.status) 
#reading copy number data
copy_number <- read_copy_number() %>%
  dplyr::select(!patient_id)

#merging clinical and copy number data to make annotation with all data (rsem_annot_all)
rsem_annot_all <- merge(rsem_annot,
                        clinical_data,
                        by = "sample_id",
                        all.x = TRUE)
rsem_annot_all <- merge(rsem_annot_all,
                        copy_number,
                        by = "sample_id",
                        all.x = TRUE)
rownames(rsem_annot_all) <- rsem_annot_all$bcr_patient_barcode
```

## re-setting column order of counts matrix to be the same as the rownames of annotation file
```{r, eval = TRUE, error = FALSE, warnings = FALSE}
#re-setting columns of matrix
rsem_long <- rsem_long[,rownames(rsem_annot_all)]

#double checking that rows and columns match between the two variables
print(all(rownames(rsem_annot_all) == colnames(rsem_long)))

#removing irrelevant variables up to this point
rm(clinical_data,
   copy_number,
   rsem,
   rsem_annot)
```

# Quality control of rnaseq raw counts & normalizing data to counts per million (CPM). Answers question 1 - "QC and normalize the RNA dataset."
## Setting up descriptive statistics of clinical data & total counts per million per sample type as a QC metric
```{r, eval = TRUE, error = FALSE, warnings = FALSE}
T_cpm <- merge(as.data.frame(colSums(rsem_long)/1000000) %>%
              dplyr::rename(total_counts_per_million = `colSums(rsem_long)/1e+06`) %>%
              mutate(bcr_patient_barcode = rownames(.)),
              rsem_annot_all,
              by = "bcr_patient_barcode",
              all.x = TRUE)
```

## Violin of total counts per million by sample type (QC metric)
```{r, eval = TRUE, error = FALSE, warnings = FALSE}
print(ggplot(T_cpm,
            aes(x=sample_type,
                y=total_counts_per_million)) + 
  geom_violin() + 
    geom_boxplot(width=0.1) + 
    theme_classic())
```

## Descriptive statistics associated with total counts per million & sample types
```{r, eval = TRUE, error = FALSE, warnings = FALSE}
print(by(T_cpm[,c("total_counts_per_million",
            "sample_type")],
   T_cpm$sample_type,
   summary))
```

## Normalizing RNA-seq data set to counts per million. TPM is more ideal, however gene length is not available and CPMs will suffice for excercize. 
```{r, eval = TRUE, error = FALSE, warnings = FALSE}
#Getting the normalization factor sum of columns/ million
normFactor <- colSums(rsem_long)/1000000

#normalizing counts matrix with normalization factor to get CPM normalized matrix
rsem_long_norm <- rsem_long/normFactor
```

# Exploring unsupervised clustering of primary tumor RNA-seq data. Answers question 3 - Cluster the cancer RNA data (PCA, UMAP, etc). 

## PCA Analysis
```{r, eval = TRUE, error = FALSE, warnings = FALSE}
#log transforming normalized matrix for PCA 
rsem_long_norm_log <- log2(rsem_long_norm+1)

#Principle Component Analysis of 915 (out of 1093) Primary Tumors with IHC status available. Not including metastasis data as there are only 7 samples and removing it will reduce sample type variability. 
her2_ihc_patients <- rsem_annot_all %>%
  filter(sample_type == "Primary Tumor") %>%
  drop_na(ihc_her2)
pca <- prcomp(t(rsem_long_norm_log[,c(rsem_annot_all[her2_ihc_patients$bcr_patient_barcode,]$bcr_patient_barcode)]))

#Retrieving PC1 and PC2 and adding back metadata for visualization
pca_df <- merge(data.frame(pca$x[,c(1,2)]),
                her2_ihc_patients,
                by = "row.names")

#Plotting unsupervised clustering of Primary Tumors overlayed with HER2 status for IHC. There is no clear clustering based on IHC status. 
a <- ggplot(data = pca_df %>%
              drop_na(ihc_her2),
       aes(x = PC1, 
           y = PC2,
           color = ihc_her2)) +
    geom_point(size = 1) +
    theme_classic()

print(a)

#removing unneeded variables
rm(pca_df)
```

## Unsupervised heigharcal clustering and heatmap vizualization
```{r, eval = TRUE, error = FALSE, warnings = FALSE}
# Calculate variance between samples
data<- rsem_long_norm_log[,her2_ihc_patients$bcr_patient_barcode]
vars <- apply(data, 1, var)

#subsetting data to 0.99% percentile
data2 <- data[vars >= quantile(vars, 0.99),]
rse_annot_all_data2 <- rsem_annot_all[colnames(data2),] %>%
  dplyr::select(ihc_her2)

# Heatmap visualizing the top 206 variable genes with her2 ihc annotation. This clearly shows that her2 positivity is associated with patients in the far right cluster. 
pheatmap(data2,
         show_rownames = FALSE,
         show_colnames = FALSE,
         cutree_cols = 4,
         annotation_col = rse_annot_all_data2)
```
